Adding Sensitivity to 21\,cm Inteferometric Probes of Reionization by Optimizing Choice of Baselines

### 2017-05-03 Rotation paper/draft_working.tex
save

\begin{abstract}
The observational effort to measure the cosmological 21\,cm power spectrum with radio interferometers requires high sensitivity. We introduce a straightforward and accurate method to extract sensitivity from visibility data. We develop our method in line with existing visibility-based power spectrum pipelines. Our method efficiently finds subsets of baselines to cross correlate, including partially redundant baselines, and quantifies the sensitivity contributions of given baselines. 
Using the 128-element configurations and beams of Precision Array for Probing the Epoch of Reionization (PAPER-128), as well as 4 planned versions of the Hydrogen Epoch of Reionization Array (HERA), we illustrate how our method applies to different arrays and predict the sensitivity improvements of including each baseline pairs. We show that inclusion of non-perfectly redundant baselines 
would account for $20\%$ to $60\%$ of the sensitivity of PAPER-128 and different configurations of HERA. 
\end{abstract}

@@

Analysis pipelines of such maximum-redundancy arrays typically fall into two categories. In the first, more traditional technique, image is formed in Fourier domain through rotation synthesis, then foreground is either removed or avoided to construct a power spectrum. The second, lately popular technique works directly with visibilities from baselines and cross multiply them to form the power spectrum. Due to its more natural focus on the instrumental output of visibilities, this technique is shown to fair better in dealing with systematics such as RFIs and instrumental cross-talks, and thus tend to provide good limits on the power spectrum. \cite{Ali2015} provides the newest upper limit to the power spectrum
measurements with the 64-element version of PAPER (henceforth referred to as PAPER-64). However, there is one caveat with existing visibility-based pipelines. While an imaging based power-spectrum pipeline naturally includes all redundancy information, visibility-based pipelines until now only cross multiply fully redundant baselines, in other words baselines of the same length and orientation.

Baselines of different length and orientation still contain partially redundant information. The use of partially redundant baselines is in principle not a fundamental limitation of visibility-based pipelines. In fact, most instrumental reports to date do include partial redundancy. Recently, \cite{wterm} proposed an algorithmic approach to extract power-spectrum from partially redundant baselines in tracking measurements, by carefully treating the $w$-term in the fringe pattern to obtain weights for cross-correlation. Our work parallels this effort by focusing on arrays of the drift-scan type that do not have tracking capabilities, such as PAPER, HERA and SKA.
The Earth's rotation causes the baselines in a drift-scan array to pick up different modes of the sky with time. Rotation-synthesis makes use of the rotation-induced $uv$ coverage map to form image. In a visibility based pipeline, we can extract the same rotation-induced redundancy. In this point of view, baselines that are slightly different in length and orientation
rotate into each other at a time delay. We can thus cross-multiply time-shifted visibilities, with the proper weighting, to form power spectrum.  Due to the large number of elements of modern arrays (upward of 100), the task of cross-multiplying every baseline against every other, scaling as number of array element to the fourth power, can be computationally formidable. Our contribution is thus twofold. First we introduce a formalism to include pairs of partially-redundant baselines in a visibility based power spectrum pipeline. Secondly we show how to use the formalism to automaticaly preselect baseline pairs and time offsets, making the problem computationally efficient. More precisely, our formalism allows one to simultaneously a) identify the baselines that give good
redundancy, b) find the time offset that corresponds to maximal redundancy for a given pair of baselines, and
c) quantify the sensitivity associated with cross multiplying
such a pair of partially redundant baselines, which in turns is used as weight to combine measurements in a power spectrum pipeline.

@@

In order to eliminate confusion and ambiguity, we
introduce some terminology for the rest of this paper, some of which differs from what is commonly found in the literature.

First, we make the distinction of a \textit{``baseline''}, which refers to two specific antennas, and a \textit{``class of baselines''}, which refers to all baselines of the same length and orientation in a given array.

Baselines of the same class are traditionally called
``redundant baselines'', because they measure the same Fourier mode
in the sky.  We shall call baselines in the same class \textit{``equivalent baselines''}, and reserve the word \textit{``redundancy''} of two baselines in reference to a variable function of the relative time-offsets of their visibility time series. With this terminology, two equivalent
baselines are fully redundant with each other simultaneously at all
times. Non-equivalent baselines also have partial redundancy, which is maximized if their
respected time series are shifted with respect to one another by some delay. In other words, one baseline can be ``rotated into''
another if they are \textit{``near-equivalent''}.

@@

\caption{The PAPER-128 layout. Each dot corresponds to the location of
an antenna. Top panel shows the antenna positions drawn to scale;
bottom panel show the antenna labels and distances, excluding the outrigger
antennas.
The numbering of the antennas the bottom panel are original labels
during instrument assembly and does not bear significant
meaning. In the bottom panel, the two baselines (5,77) and (40,85) are an example of equivalent pair, with separation denoted by \{1, 0\}, for the
antennas are separated by 1 unit east and 0 unit north. Similarly,
the baselines (5,69) and (5,85) are examples
of the classes \{1, 1\} and \{1,-1\}, respectively. Note \{1, 0\} and \{-1, 0\} for example
are the same baseline classes and should not be counted twice.}
\label{fig:AntPos}
\end{figure}

We shall use the 128-element PAPER array (henceforth as PAPER-128) to motivate our formalism and demonstrate our method, and extend our results to several HERA configurations in Sec \ref{sec:arrconf}. 
The PAPER array is located in the Karoo desert in South Africa (30:43:17.5
S, 21:25:41.8 E). The layout pattern with antenna labels are show
in Fig. \ref{fig:AntPos}. The array consists of a 112-element grid and 16 "outriggers" used primarily to aid calibration. We see the antenna spacing in North-South
directions are comparatively close (4m), so that baselines such as
(0, 44) and (0, 7) are very close to equivalent. In the bottom panel, the two baselines (5, 77) and (40, 85) are an example of equivalent pair, with separation denoted by \{1, 0\}, for the
antennas are separated by 1 unit east and 0 unit north. Similarly,
the baselines (5, 69) and (5, 85) are examples
of \{1, 1\} and \{1, -1\}, respectively.
Note \{1, 0\} and \{-1, 0\} for example are the same classes and should
not be counted twice. Antennas in purely north-south baselines
are close enough to induce cross-talk, and hence are not suitable
for use. The PAPER-64 analysis of \cite{Ali2015} used three classes of baselines, the PAPER-128
equivalent of which are 
\{2, 0\}, \{2, 1\} and \{2, -1\}. There each of these classes
of baselines are cross multiplied to itself.

>>>>

\begin{abstract}
The observational effort to measure the cosmological 21\,cm power spectrum with radio interferometers requires high sensitivity. Current visibility-based power spectrum pipelines, though shown to ease control of systematics, lack the ability to include partial redundancy. We introduce a method to include partial redundancy in such power spectrum pipelines of drift-scan arrays. Our method efficiently finds pairs of baselines to cross-multiply, and quantifies the sensitivity contributions of given baselines. 
Using the 128-element configurations and beams of the Precision Array for Probing the Epoch of Reionization (PAPER-128), as well as 4 planned versions of the Hydrogen Epoch of Reionization Array (HERA), we illustrate how our method applies to different arrays and predict the sensitivity improvements of including each baseline pairs. We show that inclusion of partially redundant baselines 
would account for $20\%$ to $60\%$ of the sensitivity of PAPER-128 and different configurations of HERA. 
\end{abstract}

@@

Analysis pipelines of such maximum-redundancy arrays typically fall into two categories. In the first, more traditional technique, image is formed in Fourier domain through rotation synthesis, then foreground is either removed or avoided to construct a power spectrum. The second, lately popular technique works directly with visibilities from baselines and cross multiply them to form the power spectrum. Due to its more natural focus on the instrumental output of visibilities, this technique is shown to fair better in dealing with systematics such as RFIs and instrumental cross-talks, and thus tend to provide good limits on the power spectrum. \cite{Ali2015} provides the newest upper limit to the power spectrum
measurements with the 64-element version of PAPER (henceforth referred to as PAPER-64). However, one disadvantage of existing visibility-based pipelines is the lack of use of partial redundancy. Baselines of different length and orientation still contain partially redundant information. While an imaging based power-spectrum pipeline naturally includes all redundancy information, visibility-based pipelines until now only cross multiply fully redundant baselines, in other words baselines of the same length and orientation.

The use of partially redundant baselines is in principle not a fundamental limitation of visibility-based pipelines. In fact, most instrumental reports to date do include partial redundancy. Recently, \cite{wterm} proposed a visibility-based approach to extract power spectrum from partially redundant baselines in tracking measurements. Our work parallels this effort by focusing on arrays of the drift-scan type that do not have tracking capabilities, such as PAPER, HERA and SKA.
The Earth's rotation causes the baselines in a drift-scan array to pick up different modes of the sky with time. Rotation-synthesis makes use of the rotation-induced $uv$ coverage map to form image. In a visibility based pipeline, we can extract the same rotation-induced redundancy. In this point of view, baselines that are slightly different in length and orientation
``rotate into'' each other at a time delay. We can thus cross-multiply time-shifted visibilities, with the proper weighting, to form power spectrum.  Due to the large number of elements of modern arrays (upward of 100), the task of cross-multiplying every baseline against every other, scaling as number of array element to the fourth power, can be computationally formidable, and many pairs of baselines provide only negligible redundancy information. Our contribution is thus twofold. First we introduce a formalism to include pairs of partially-redundant baselines in a visibility based power spectrum pipeline. Secondly we show how to use the formalism to automatically pre-select baseline pairs and time offsets, making the problem computationally efficient. More precisely, our formalism allows one to simultaneously a) identify the baselines that give good
redundancy, b) find the time offset that corresponds to maximal redundancy for a given pair of baselines, and
c) quantify the sensitivity associated with cross multiplying
such a pair of partially redundant baselines, which in turns is used as weight to combine measurements in a power spectrum pipeline.

@@

In order to eliminate confusion and ambiguity for the rest of this paper, we
introduce some terminology that may differ from what is commonly found in the literature.

We make the distinction of a \textit{baseline}, which corresponds to two specific antennas, and a \textit{class of baselines}, which refers to all baselines of the same length and orientation in a given array. 
Baselines of the same class are traditionally called
``redundant baselines'', because they measure the same Fourier mode
in the sky.  We shall call baselines in the same class \textit{equivalent baselines}, and reserve the word \textit{redundancy} of two baselines in reference to a variable function of the relative time-offsets of their visibility time series. With this terminology, two equivalent
baselines are fully redundant with each other simultaneously at all
times. Non-equivalent baselines also have partial redundancy, and the redundancy can be maximized if their
respected time series are shifted with respect to one another by some delay. In other words, one baseline can be ``rotated into''
another if they are \textit{near-equivalent}.

@@

\caption{The PAPER-128 layout. Each dot corresponds to the location of
an antenna. Top panel shows the antenna positions drawn to scale;
bottom panel show the antenna labels and distances of the 112-element grid, in other words excluding the outrigger antennas.
The numbering of the antennas in the bottom panel are original labels
during instrument assembly and does not bear significant
meaning. In the bottom panel, the two baselines (5,77) and (40,85) marked with red segments are example of an equivalent pair, both with separation denoted by \{1, 0\}, for the
antennas are separated by 1 unit east and 0 unit north. Similarly,
the baselines marked in blue and yellow are examples
of the classes \{1, 1\} and \{1,-1\}, respectively. Due to the small North-South separation within the grid, \{1, 1\} is expected to be near-equivalent to \{1, 0\}. }
\label{fig:AntPos}
\end{figure}

We shall use the 128-element PAPER array (henceforth as PAPER-128) to motivate our formalism and demonstrate our method, and extend our results to several HERA configurations in Sec \ref{sec:arrconf}. 
The PAPER array is located in the Karoo desert in South Africa (30:43:17.5
S, 21:25:41.8 E). The layout pattern with antenna labels are show
in Fig. \ref{fig:AntPos}. The array consists of a 112-element grid and 16 "outriggers" used primarily to aid calibration.  In the bottom panel, the two baselines (5, 77) and (40, 85), marked in red, are an example of equivalent pair. We denote a equivalency-class of baselines in the PAPER grid by their separations, in this case  \{1, 0\}, for the
antennas are separated by 1 unit east and 0 unit north. Similarly,
the baselines marked in blue and yellow are respectively examples
of \{1, 1\} and \{1, -1\}.
Note \{1, 0\} and \{-1, 0\} for example are the same class and should
not be counted twice. Antennas in purely north-south baselines
are close (4m), and hence these baselines are not suitable
for cross-multiplication due to cross-talks. On the other hand, the close North-South separation means that classes such as \{1, 0\} and  \{1, 1\} are expected to be near-equivalent. The PAPER-64 analysis of \cite{Ali2015} used three classes of baselines, the PAPER-128
equivalent of which are 
\{2, 0\}, \{2, 1\} and \{2, -1\}. There each of these classes
of baselines are cross multiplied within itself. This paper provides the method for inter-class multiplications.
